Working from home and digital divides: resilience in the time of a pandemic

### 2020-12-24 paper/v1_sage/broadband.speed.covid.Rmd
conclusions

---
title: WFH and broadband speed (title needs rework)
author:
- name: A. Uthor*
  num: 1,2
- name: O. Tro
  num: 2
- name: O. Vriga
  num: 3
address:
- num: 1
  org: Department of Incredible Research, University A, City A, Country A
- num: 2
  org: Department of Applied Things, University B, City B, Country B
- num: 3
  org: Very Important Stuff Committee, Institute C, City C, Country C
corrauth: "Corresponding author name, This is sample corresponding address."
email: correspondingauthor@protonmail.com
abstract: "TBC"
keywords:   covid; internet; working from home; broadband speed; time-series clusters;
classoption:
  - Royal
  - times
bibliography: bibliography
bibliographystyle: sageh
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{dcolumn}
  \def\tightlist{}
  \newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds,
may reinforce or redress the spatial and social dimensions of the digital division
exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
Our analysis enables us to better understand how the spatial and social distribution
of both occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era,
and broadband services and infrastructure must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used
interchangeably, as most remote labour during the Covid-19 crisis was carried out
in the homes of individual employees rather than any other location [@eurofound2020].
However, it should be noted that previous research has explored how telecommuting
can occur in other places, including satellite offices or on public transport
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of
telecommuting within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or part-time, and
those who use digital technologies to work remotely full-days or part-days
[@allen2015effective; @bailey2002review; @haddad2009examination].
No matter the definition, the option and capability to telecommute or work from home
has never been equally distributed spatially or socio-economically any more than
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK found
that telecommuters are most likely to hold professional, managerial, and technical
occupations where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].

@@

Based on the outcomes of the time-series clustering, we identify $J$ distinct
and crisp clusters.
We then regress this cluster membership against a vector $X_{i}$ of
socio-economic and geographic variables, which are discussed in detail in
the relevant Section [4.2]](#sec:4.2), in order to explore how the different patterns
might support or undermine efforts to work from home and maintain safe productivity
and whether they reinforce existing spatial and social inequalities.
This analysis enables us to provide a more nuanced understanding of how
telecommuting and technology intersect at a time of extreme demand, and what
lessons this time has for a future where telecommuting is likely to remain a
common means of accessing work and broadband services, as well as infrastructure,
must be fit for purpose.

@@

The temporal profiles of the local authority clusters have been summarised in
Figures \ref{UpClusterL} and \ref{UpClusterS} and Table \ref{up.cluster.descr}.
The graphs show a composite profile of mean upload speeds per hour
per day for each cluster, with the largest, in terms of the LAD membership and
population, five clusters in Figure \ref{UpClusterL},
and the next six in Figure \ref{UpClusterS}.
These figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK,
the temporal clusters offering a novel approach to understanding spatial disparity.

@@

stargazer(clustersLADup[1:7],
          summary = F,
          font.size = "footnotesize",
          type = "latex",
          header = F,
          digit.separator = ",",
          digit.separate = c(3,3),
          covariate.labels=c("Cluster",
                             "N. of LADs",
                             "LAD population",
                             "mean speed",
                             "SD speed",
                             "mean AM speed",
                             "mean PM speed"),
                             #"AM_slower",
                             #"SDtoMean"),
          rownames = F,
          no.space=TRUE,
          column.sep.width = "0pt",
          notes = "Note: All speed measures are upload speeds",
          title = "Upload speed cluster characteristics\\label{up.cluster.descr}")
```
The second largest cluster, comprising $126$ local authorities and over $20$ million
people, is cluster $13$.
Cluster $13$ has the slowest aggregate mean upload speed of any of the clusters,
and the second highest ratio of the standard deviation to the mean.
This suggests that those living in local authorities in this cluster experienced
some of the lowest quality broadband services in terms of upload speeds in the UK.
However, the variation in upload speeds in cluster $13$,
which can be an indication of its reliability, does not seem to disproportionately
affect the morning peak from $9:00$-$10:59$, as upload speeds are, on average,
only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$,
when entertainment purposes are likely to be using the most bandwidth.
In comparison, the five LADs that are home to $343$ thousand people in cluster $1$
not only experience the second slowest mean upload speeds
and the highest ratio of standard deviation to the mean,
but are also much more affected during the morning peak.

Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs home to $29.5$
million people, experience aggregate mean upload speeds of about $1$Mb/s faster
than those in cluster $13$, but still lower than the other three large clusters
and most of the smaller clusters, suggesting a middling quality of service.
However, the time profile for cluster $6$ in Figure \ref{UpClusterL} shows
that upload speeds are highest at $6:00$ on a Monday morning
and peak between $23:00$ and midnight on Wednesday and Thursday,
but tend to be lower during the working day.
Indeed, experienced mean upload speeds in the morning peak are $4.4$% lower
than in the evening peak -- a greater, more noticeable change than any of the other
large clusters experience, but smaller than any of the clusters with temporal
profiles shown in Figure \ref{UpClusterS}.
Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
clusters $1$ and $13$, and their standard deviation is not dissimilar.
However, this masks great variation in when lower speeds are experienced,
with the mean upload speeds much lower between $9:00-10:59$ than
between $19:00-20:59$ in cluster $8$, but slightly faster in the morning in
cluster $12$.

@@

Using auxiliary regressions, we test whether the clusters that have higher mean
speeds and more reliable services consist of LADs that are more urban, affluent,
and / or more likely
to benefit from a choice of high quality internet services.
We also estimate which of our clusters are more likely to have a higher proportion of
occupations which enable telecommuting because of the nature of the work.
The results of these auxiliary regressions are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
[methods and data](#sec:3) section
and equation \ref{eq1} and each column represent a different cluster.
The reference case is cluster $4$, which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
Mean, experienced upload speeds in cluster $4$ are close to both the average speeds
for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
($9.3$Mb/s) as shown in Table \ref{up.cluster.descr}.
However, the standard deviation for cluster $4$ and the difference between average
speeds in the morning compared to the evening peak periods are indications of
worse reliability than many of the other clusters.
Hence, the results in Table \ref{aux} should be seen as relative
rather than absolute probabilities.

First, the number of speed tests run per cluster inhabitant between $9:00$-$10:59$
is an indication of satisfaction or at least a lack of concern over broadband speeds
and quality of service.
People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests
(at a per capita basis)
at this time of day, whilst people in cluster $11$ were the least likely to do so.
This high level of satisfaction matches the high quality and reliability of service
we would expect from the analysis in Section [4.1]](#sec:4.1).
Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per
capita than those in the other six multi-authority clusters.
Cluster $7$ was the only cluster with higher mean speeds and middle to high measures
of reliability which was more likely to run more speed tests per capita.
This might be a reflection of the choice of ISP and connection options in those
LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have
Virgin Media connections.
Fast Virgin Media cable connections have historically only been available
to $45$% of premises in the UK [@ofcom2016], where the more lucrative and
competitive market originally attracted the cable TV provider.
Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
Virgin connections, which is an indication that people in these clusters are
more likely to live in urban areas, with more choice of broadband services,
or in other words, are more likely to be on the right side of the first level,
infrastructure-based digital divide.

We employ distance from London and for the nearest metropolitan area (including
London) as two more variables depicting peripherality, urban structure and,
potentially, first level digital divides.
The broadband speed tests run in the authorities in cluster $3$ are more likely to
be taking place close to London than those run in any of the other clusters,
a result that makes sense considering two of the four authorities in cluster $3$
are the London commuter towns of Harlow and Luton.
However, even though London was also included in the variable calculating distance
from the centre of one of either the ten largest metropolitan areas in England,
or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
Still, whilst significant, the effect is tiny and the other members of cluster $3$
are Corby and Eastbourne, two towns which, whilst not part of metropolitan areas,
are also not in remote parts of the country, and are home to an estimated population
of $175$ thousand people between them. **move maps to here?**

Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
metropolitan area, even though the four local authorities of cluster $7$ include
no central urban boroughs and only one LAD that is part of a metropolitan area of
governance - Tameside in Greater Manchester.
This may be why those in cluster $7$ are not served by Virgin Media.
It also is a demonstration of the complexity of experienced broadband upload speeds
as captured by time-series clustering, and their likely interaction with first level
digital divides.
Cluster $1$, which our analysis suggests lacks broadband resilience, contains five,
mainly rural authorities.
However, they are scattered around the country --
Ceredigion in West Wales, Darlington in the Northeast, Eden between the Lake District
and the North Penines, Rossendale in the Northwest, and Rutland in the East Midlands
-- and therefore, the results in Table \ref{aux} indicate that these authorities
are not as far from the centre of a metropolitan area as authorities in cluster $3$,
but furthest from London.

@@

Yet we need to consider the results for other variables in Table \ref{aux} to better
determine whether clusters $1$ and $13$, which appear to suffer most from a lack 
of quality internet services, are also more likely to have a low skilled workforce, 
less able to benefit from telecommuting. 
Cluster $1$ has the lowest propotion of educated people of any of the clusters,
and the lowest earnings, despite recording the highest proportion of managerial, 
professional jobs, and the second highest proportion of tech jobs.
Rural areas such as those in cluster $1$ are often home to many retired people,
which might explain these results or perhaps, as these figures are relative,
we could note that cluster $1$ also has a greater proportion of skilled trades
than other clusters.
Either way, it seems likely that the first level digital divide reinforces 
other inequities in cluster $1$.
In comparison, those cluster $13$ are more likely to earn more and have a better
education.
Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the 
middle in terms of job density.
If this is an indication of a high number of SMEs, it might give some explanation
for the variable internet quality, considering that small businesses have not been
seen as the most valuable customers for higher speed broadband services **insert ref**.

Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more 
noise within the averages we use to measure internet quality and reliability,
and the result indicates an average level of service, even if better than LADs
in cluster $13$.
Cluster $6$ also falls towards the middle of the clusters on many of the socio-
economic variables in Table \ref{aux}. 
It ranks third or fourth out of the eleven clusters in the table in terms of
having a higher proportion of managerial, professional, and tech jobs, as well as 
higher earnings, but is fourth from bottom in terms of education.
The LADs in cluster $6$ are also diverse, with few truly remote areas, 
but many urban areas in both the North, including Birmingham, Leeds and Sheffield, 
and South, including twelve London Boroughs, and many suburban areas and 
smaller cities like Oxford and Cambridge.
The capital cities of the other UK nations, Belfast, Cardiff and Edinburgh, are
also in this cluster, suggesting perhaps that the lower level of reliability 
discussed in Section [4.1](#sec:4.1) may be as a result of increased demand,
e.g. for telecommuting.

In comparison, clusters $9$ and $11$, which benefit from the most resilient
internet connections, are characterised by lower share of occupations that are  
may benefit from the ability to telecommute (e.g. managerial and technical). 
Cluster $9$ is more highly educated and higher earning than cluster $11$, 
has fewer businesses per inhabitant and less job density.
Indeed, cluster $11$ has a negative coefficient for the NVQ4 variable, whilst
for cluster $9$ the coefficient is positive, and cluster $11$ has the second
highest job density and cluster $9$ the second lowest. 
The former might be an indication of the higher returns that resilient
broadband infrastructure generates for cluster $9$ versus cluster $11$. 
Indeed, we did detect slightly more slowdown in the AM peak in cluster $9$ than in
cluster $11$ (see Table \ref{up.cluster.descr}).
Meanwhile, a propensity for a high job density might indicate more labour intensive 
economic structures in cluster $11$ than in the LADs in cluster $9$.

The LADs in clusters $9$ and $11$ also have opposite signs for their likelihood 
of being in the South of England.
Those in cluster $9$ less likely to be in the South and those in cluster $11$ 
more likely, although the coefficient for cluster $9$ might be skewed by 
the presence of the Scottish cities of Glasgow and Aberdeen, 
as Scotland has a different economic profile than England.
Still, both clusters consist of larger LADs in terms of population, 
including the urban centre of Manchester and three of the nine other districts of
Greater Manchester, Nottingham, both Portsmouth and Southampton,
eight London Boroughs, and four of the seven constituent authorities of the
West Midlands Combined Authority.
Among the $65$ LADs are also a number of other tightly bounded urban areas, 
such as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas 
at the centre of less confined districts like Burnley, 
the digital hub of Milton Keynes, and Northampton.
Thus, our findings confirm the importance of geography and urban locations 
for being on the right side of the first level digital divide, and suggest that 
internet resilience cuts across other digital and economic divides.

@@

In comparison, LADs in cluster $12$ are less likely to be able to benefit from
quality internet services if they had them, with fewer individuals achieving 
NVQ4 or better and lower levels of occupations that would benefit from homeworking. 
Yet LADs in this cluster benefit from the second highest level of earnings, and
have the largest stock of businesses per inhabitant and the highest job density. 
This density of businesses could be why cluster $12$ is the only cluster with higher
speeds in the morning peak during the study period.
If people are at home, business premises might well be abandoned.
Cluster $12$ is home to $1.5$million people spread across twelve LADs, including
the City of Westminster and the London Borough of Hammersmith.
However, it also includes a LAD in Northern Ireland, two in Wales, two suburban areas
North of London, Preston in the Northwest, East Staffordshire in the Midlands, and
Fenland in the East of England.
This spatial diversity demonstrates that the temporal clustering of internet 
resilience does not create spatial clusters, and digital divides do not necessarily
overlap with economic ones.

@@

Upload speeds are not used as benchmarks in the UK, nor have been seen as integral to
universal service, considering there has never before been such extreme demand for
telecommuting and operations such as video calls.

@@

<!-- WFH broader consequences-->
The long-term effects of such drastic changes in telecommuting and attitudes
towards working from home are difficult to predict. Nevertheless,
they span through various aspects of economy and society: from changes to
transportation planning due to altered commuting patters, to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713]**also 2020 Swedish article from JTG**; and from productivity 
and innovation changes, to changes
in agglomeration externalities and the attraction of large cities [@econobs]
just to name a few. This paper is positioned to support endeavours in
understanding the effects of increased telecommuting by
exposing the spatial and social dimensions of telecommuting including the
resilience of broadband speeds in terms of both quality and reliability of
service, and whether this reinforces or redresses prior digital divisions.
**took this bit you wrote to put in the discussion at the end?**

@@

for.descr <- data %>%
  dplyr::select(pop, job.dens2018, distMet,
                distLondon, south, 
                managers, tech, skilled, prof,
                admin, caring, plant, 
                earnings, nvq4,
                virgin,
                total.busi, LAD.tests.am20, pop) %>%
  mutate('total.busi/pop' = total.busi/pop,
         'LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
  dplyr::select(-total.busi, -pop, -LAD.tests.am20)

>>>>

---
title: "Working from home and digital divides: resilience in the time of a pandemic"
author:
- name: A. Uthor*
  num: 1,2
- name: O. Tro
  num: 2
- name: O. Vriga
  num: 3
address:
- num: 1
  org: Department of Incredible Research, University A, City A, Country A
- num: 2
  org: Department of Applied Things, University B, City B, Country B
- num: 3
  org: Very Important Stuff Committee, Institute C, City C, Country C
corrauth: "Corresponding author name, This is sample corresponding address."
email: correspondingauthor@protonmail.com
abstract: "TBC"
keywords:   covid; internet; working from home; broadband speed; time-series clusters;
classoption:
  - Royal
  - times
bibliography: bibliography
bibliographystyle: sageh
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{dcolumn}
  \def\tightlist{}
  \newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds,
may reinforce or redress the spatial and social dimensions of the digital division
exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK.
Our analysis enables us to better understand how the spatial and social distribution
of both occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a more common means of accessing work,
at least in comparison to the pre-Covid era,
and broadband services and infrastructure must be fit for purpose.

@@

<!-- Who works from home-->
In this analysis, the terms ‘telecommuting’ and ‘working from home^
[See also the popular social media hashtag #WFM]’ are used
interchangeably, as most remote labour during the Covid-19 crisis was carried out
in the homes of individual employees rather than any other location [@eurofound2020].
However, it should be noted that previous research has explored how telecommuting
can occur in other places, including satellite offices or on public transport
[@felstead2012rapid; @siha2006telecommuting].
Previous research has also used a variety of definitions to measure the level of
telecommuting within different workforces, distinguishing, for example, between those
directly employed, indirectly employed, self-employed, full-time or part-time, and
those who use digital technologies to work remotely full-days or part-days
[@allen2015effective; @bailey2002review; @haddad2009examination].
No matter the definition, the option and capability to telecommute or work from home
has never been equally distributed spatially or socio-economically any more than
different industries and employment opportunities have.
For example, studies from the United States, the Netherlands, and the UK found
that telecommuters are most likely to hold professional, managerial, and technical
occupations where the workforce is better educated and wealthier, and that there is
suppressed demand among women and part-time workers
[@headicar2016move; @peters2004employees; @singh2013modeling].

@@

Based on the outcomes of the time-series clustering, we identify $J$ distinct
and crisp clusters.
We then regress this cluster membership against a vector $X_{i}$ of
socio-economic and geographic variables, which are discussed in detail in
the relevant Section [4.2](#sec:4.2), in order to explore how the different patterns
might support or undermine efforts to work from home and maintain safe productivity
and whether they reinforce existing spatial and social inequalities.
This analysis enables us to provide a more nuanced understanding of how
telecommuting and technology intersect at a time of extreme demand, and what
lessons this time has for a future where telecommuting is likely to remain a
common means of accessing work and broadband services, as well as infrastructure,
must be fit for purpose.

<!-- limitations-->
Our analysis is not free of limitations as, for instance, although we do have
individual data we cannot identify individuals or households and, therefore, we
need to aggregate our data at the LAD level. However, such individual data could
only be found though surveys which by definition result in limited samples
and, importantly, less detailed data about the quality of internet connectivity.

@@

The temporal profiles of the local authority clusters have been summarised in
Figures \ref{UpClusterL} and \ref{UpClusterS} and Table \ref{up.cluster.descr}.
The graphs show a composite profile of mean upload speeds per hour
per day for each cluster, with the largest, in terms of the LAD membership and
population, five clusters in Figure \ref{UpClusterL},
and the next six in Figure \ref{UpClusterS}.
These figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK.

@@

stargazer(clustersLADup[1:7],
          summary = F,
          font.size = "footnotesize",
          type = "latex",
          header = F,
          digit.separator = ",",
          digit.separate = c(3,3),
          covariate.labels=c("Cluster",
                             "N. of LADs",
                             "LAD population",
                             "mean speed",
                             "SD speed",
                             "mean AM speed",
                             "mean PM speed"),
                             #"AM_slower",
                             #"SDtoMean"),
          rownames = F,
          no.space=TRUE,
          column.sep.width = "0pt",
          notes = "Note: All speed measures are upload speeds",
          title = "Upload speed cluster characteristics\\label{up.cluster.descr}")
```
The second largest cluster, comprising $126$ local authorities and over $20$ million
people, is cluster $13$, which has the slowest aggregate mean upload speed
of any of the clusters,
and the second highest ratio of the standard deviation to the mean.
This suggests that those living in local authorities in this cluster experienced
some of the lowest quality broadband services in terms of upload speeds in the UK.
However, the variation in upload speeds in cluster $13$,
which can be an indication of its reliability, does not seem to disproportionately
affect the morning peak from $9:00$-$10:59$, as upload speeds are, on average,
only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$,
when entertainment purposes are likely to be using the most bandwidth.
In comparison, the five LADs that are home to $343$ thousand people in cluster $1$
not only experience the second slowest mean upload speeds
and the highest ratio of standard deviation to the mean,
but are also much more affected during the morning peak.

Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs home to $29.5$
million people, experienced aggregate mean upload speeds of about $1$Mb/s faster
than those in cluster $13$, but still lower than the other three large clusters
and most of the smaller clusters, suggesting a middling quality of service.
However, the temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows
that upload speeds are highest at $6:00$ on a Monday morning
and peak between $23:00$ and midnight on Wednesday and Thursday,
but tend to be lower during the working day.
Indeed, experienced mean upload speeds in the morning peak are $4.4$% lower
than in the evening peak -- a greater, more noticeable change than any of the other
large clusters experience, but smaller than any of the clusters with temporal
profiles shown in Figure \ref{UpClusterS}.
Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
clusters $1$ and $13$, and their standard deviation is not dissimilar.
However, this masks great variation in when lower speeds are experienced,
with the mean upload speeds much lower between $9:00-10:59$ than
between $19:00-20:59$ in cluster $8$, but slightly faster in the morning in
cluster $12$.

@@

Using an auxiliary multinomial regression, we test whether the clusters that
have higher mean
speeds and more reliable services consist of LADs that are more urban, affluent,
and / or more likely
to benefit from a choice of high quality internet services.
We also estimate which of our clusters are more likely to have a higher proportion of
occupations which enable telecommuting because of the nature of the work.
The results of the auxiliary regression are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
[methods and data](#sec:3) section
and equation \ref{eq1} and each column represent a different cluster.
The reference case is cluster $4$, which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
Mean, experienced upload speeds in cluster $4$ are close to both the average speeds
for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
($9.3$Mb/s) as shown in Table \ref{up.cluster.descr}.
However, the standard deviation for cluster $4$ and the difference between average
speeds in the morning compared to the evening peak periods are indications of
worse reliability than many of the other clusters.
Hence, the results in Table \ref{aux} should be seen as relative
rather than absolute probabilities.

First, we control for the number of speed tests run per cluster inhabitant
between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
connections. Regarding the former, we expect people in LADs with more unreliable
connections to test more often their internet speeds and the results validate our
priors. Regarding the former, fast Virgin Media cable connections have historically
only been available to $45$% of premises in the UK [@ofcom2016], where the more
lucrative and competitive market originally attracted the cable TV provider.
Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
Virgin connections, which is an indication that people in these clusters are
more likely to live in urban areas, with more choice of broadband services,
or in other words, are more likely to be on the right side of the first level,
infrastructure-based digital divide.

<!-- is an indication of satisfaction or at least a lack of concern over broadband speeds -->
<!-- and quality of service. -->
<!-- People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests -->
<!-- (at a per capita basis) -->
<!-- at this time of day, whilst people in cluster $11$ were the least likely to do so. -->
<!-- This high level of satisfaction matches the high quality and reliability of service -->
<!-- we would expect from the analysis in Section [4.1](#sec:4.1). -->
<!-- Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per -->
<!-- capita than those in the other six multi-authority clusters. -->
<!-- Cluster $7$ was the only cluster with higher mean speeds and middle to high measures -->
<!-- of reliability which was more likely to run more speed tests per capita. -->
<!-- This might be a reflection of the choice of ISP and connection options in those -->
<!-- LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have -->
<!-- Virgin Media connections. -->

Then, we employ distance from London and for the nearest metropolitan area (including
London) as two variables depicting peripherality, urban structure and,
potentially, first level digital divides.
The broadband speed tests run in the authorities in cluster $3$ are more likely to
be taking place close to London than those run in any of the other clusters,
a result that makes sense considering two of the four authorities in cluster $3$
are the London commuter towns of Harlow and Luton.
However, even though London was also included in the variable calculating distance
from the centre of one of either the ten largest metropolitan areas in England,
or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
Still, whilst significant, the effect is tiny and the other members of cluster $3$
are Corby and Eastbourne, two towns which, whilst not part of metropolitan areas,
are also not in remote parts of the country, and are home to an estimated population
of $175$ thousand people between them.

Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
metropolitan area, even though the four local authorities of cluster $7$ include
no central urban boroughs and only one LAD that is part of a metropolitan area of
governance - Tameside in Greater Manchester.
This may explain why those in cluster $7$ are not served by Virgin Media.
It also is a demonstration of the complexity of experienced broadband upload speeds
as captured by time-series clustering, and their likely interaction with first level
digital divides.
Cluster $1$, which our analysis suggests lacks broadband resilience, contains five,
mainly rural authorities.
However, they are scattered around the country --
Ceredigion in West Wales, Darlington in the Northeast, Eden between the Lake District
and the North Penines, Rossendale in the Northwest, and Rutland in the East Midlands
-- and therefore, the results in Table \ref{aux} indicate that these authorities
are not as far from the centre of a metropolitan area as authorities in cluster $3$,
but furthest from London.

@@

Yet we need to consider the results for other variables in Table \ref{aux} to better
determine whether clusters $1$ and $13$, which appear to suffer most from a lack
of quality internet services, are also more likely to have a low skilled workforce,
less able to benefit from telecommuting.
Cluster $1$ has the lowest propotion of educated people of any of the clusters,
and the lowest earnings, despite recording the highest proportion of managerial,
professional jobs, and the second highest proportion of tech jobs.
Rural areas such as those in cluster $1$ are often home to many retired people,
which might explain these results or perhaps, as these figures are relative,
we could note that cluster $1$ also has a greater proportion of skilled trades
than other clusters.
Either way, it seems likely that the first level digital divide reinforces
other inequities in cluster $1$.
In comparison, those in cluster $13$ are more likely to earn more and have a better
education.
Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
middle in terms of job density.
If this is an indication of a high number of SMEs, it can explain
the variable internet quality, considering that small businesses have not been
seen as the most valuable customers for higher speed broadband services **insert ref**.

Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more
noise within the averages we use to measure internet quality and reliability,
and the result indicates an average level of service, even if better than LADs
in cluster $13$.
Cluster $6$ also falls towards the middle of the clusters on many of the socio-
economic variables in Table \ref{aux}.
It ranks third or fourth out of the eleven clusters in the table in terms of the
likelihood of
having a higher proportion of managerial, professional, and tech jobs, as well as
higher earnings, but is fourth from bottom in terms of education.
The LADs in cluster $6$ are also diverse, with few truly remote areas,
but many urban areas in both the North, including Birmingham, Leeds and Sheffield,
and South, including twelve London Boroughs, and many suburban areas and
smaller cities like Oxford and Cambridge.
The capital cities of the other UK nations, Belfast, Cardiff and Edinburgh, are
also in this cluster, suggesting perhaps that the lower level of reliability
discussed in Section [4.1](#sec:4.1) may be as a result of increased demand,
e.g. for telecommuting.

In comparison, clusters $9$ and $11$, which enjoy the most resilient
internet connections, are characterised by higher share of occupations that
may benefit from the ability to telecommute (e.g. managerial).
Also, LADs in Cluster $9$ are more likely to host highly educated individuals with
higher earning than cluster $11$, which has fewer businesses per inhabitant and
lower job density.
Indeed, cluster $11$ has a negative coefficient for the NVQ4 variable, whilst
for cluster $9$ the coefficient is positive, and cluster $11$ has the second
highest job density and cluster $9$ the second lowest.
The former might be an indication of the higher returns that resilient
broadband infrastructure can generate for cluster $9$ versus cluster $11$.
Indeed, we did detect slightly more slowdown in the AM peak in cluster $9$ than
in cluster $11$ (see Table \ref{up.cluster.descr}).
Meanwhile, a propensity for a higher job density indicate higher urbanization
rate in LADs in cluster $11$ than those in cluster $9$.

LADs in cluster $9$ are also less likely to be in the South and those in cluster $11$
more likely, although the coefficient for cluster $9$ might be skewed by
the presence of the Scottish cities of Glasgow and Aberdeen,
as Scotland has a different economic profile than England.
Still, both clusters consist of larger LADs in terms of population,
including the urban centre of Manchester and three of the nine other districts of
Greater Manchester, Nottingham, both Portsmouth and Southampton,
eight London Boroughs, and four of the seven constituent authorities of the
West Midlands Combined Authority.
Among the $65$ LADs are also a number of other tightly bounded urban areas,
such as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas
at the centre of less confined districts like Burnley,
the digital hub of Milton Keynes, and Northampton.
Thus, our findings confirm the importance of geography and urban locations
for being on the right side of the first level digital divide, and suggest that
internet resilience cuts across other digital and economic divides.

@@

In comparison, LADs in cluster $12$ are less likely to be able to benefit from
quality internet services if they had them, with fewer individuals achieving
NVQ4 or better and lower levels of occupations that would benefit from homeworking.
Yet LADs in this cluster benefit from the second highest level of earnings, and
have the largest stock of businesses per inhabitant and the highest job density.
This density of businesses could be why cluster $12$ is the only cluster with higher
speeds in the morning peak during the study period.
If people are at home, business premises might well be abandoned.
Cluster $12$ is home to $1.5$million people spread across twelve LADs, including
the City of Westminster and the London Borough of Hammersmith.
However, it also includes a LAD in Northern Ireland, two in Wales, two suburban areas
North of London, Preston in the Northwest, East Staffordshire in the Midlands, and
Fenland in the East of England.
This spatial diversity demonstrates that the temporal clustering of internet
resilience is not necessarily spatially dependent, and digital divides do not
necessarily overlap with economic ones.

@@

This paper offers a new perspective on telecommuting from the viewpoint of the
complex web of digital divides. Good and reliable internet connections are
necessary for the population to be able to work from home, but at the same time
not every place hosts individuals in occupations which allow for telecommuting and
not every place hosts individuals with the necessary skills to effectively use
the internet and work form home. Importantly, not every place can utilise good
internet connectivity -- even if they have them -- in order to achieve economic
resilience in a period like the current pandemic, but also for a future when
working from home might be a more common activity.

Contrary to the previous literature, we employ experienced upload speeds as a
benchmark of internet resilience concerning the capacity to work from home.
Upload speeds are not used as benchmarks in the UK, nor have been seen as integral
to universal service, considering there has never before been such extreme demand
for telecommuting and operations such as video calls. While download speeds are
associated more with video streaming services, fast upload internet speed are
necessary for seamless video calls and uploading large files, activities associated
with work. Given the granular nature of our data, we are able to consider upload
internet speeds during working hours. To analyse these data we employ time series
clustering methods, a family of unsupervised machine learning techniques which
are not often utilised in geographical research.

Using the UK as a case study, our analysis illustrated a nuanced picture, which
is the outcome of the intersection between the different digital divides and the
potential to telecommute. While some LADs appear to benefit both from reliable
internet connections and population which can work from home and, in essence,
capitalise the digital infrastructure, other LADs are suffering by both unreliable
internet and low share of individuals who can potentially telecommute.

- different places have different levels of preparedness, some places have some ground to cover

- benefits

@@

<!-- WFH broader consequences-->
The long-term effects of such drastic changes in telecommuting and attitudes
towards working from home are difficult to predict. Nevertheless,
they span through various aspects of economy and society: from changes to
transportation planning due to altered commuting patters, to changes in land
use and urban planning to accommodate people who work from home
[@BUDNITZ2020102713; @ELLDER2020102777] and from productivity
and innovation changes, to changes in agglomeration externalities and the
attraction of large cities [@econobs] just to name a few.
The take home message from this paper is that working form home and, consequently,
its long term effects cannot be understood, set aside plan our cities for, without
considering the underpinning digital divides and how telecommuting intersects
with them.

@@

for.descr <- data %>%
  dplyr::select(pop, job.dens2018, distMet,
                distLondon, south,
                managers, tech, skilled, prof,
                admin, caring, plant,
                earnings, nvq4,
                virgin,
                total.busi, LAD.tests.am20, pop) %>%
  mutate('total.busi/pop' = total.busi/pop,
         'LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
  dplyr::select(-total.busi, -LAD.tests.am20)
